Brief Report: efficacy of COVID-19 vaccination in individuals designated as clinically extremely vulnerable in Scotland

### 2021-08-29 vax4.Rmd
reinfections

---
title: "Brief Report: efficacy of vaccination in those eligible for shielding in Scotland"
author:
  - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
    affiliation: HPS
  - name: David A McAllister\textsuperscript{\getAff{Glasgow}}  
    affiliation: HPS 
  - name: Jen Bishop
    affiliation: HPS
  - name: Sharon Hutchinson\textsuperscript{\getAff{GCU}} 
    affiliation: HPS
  - name: Chris Robertson\textsuperscript{\getAff{Strathclyde}}   
    affiliation: HPS
  - name: Nazir Lone
    affiliation: Usher
  - name: Jim McMenamin 
    affiliation: HPS
  - name: David Goldberg
    affiliation: HPS
  - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
    corresponding: yes
    email: helen.colhoun@igmm.ed.ac.uk
    affiliation: HPS
address: 
- code: Usher
  address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care  
- code: HPS
  address: Public Health Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
- code: Glasgow
  address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
- code: GCU
  address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
- code: Strathclyde
  address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
- code: IGMM
  address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - AXA Chair in Medical Informatics and Epidemiology

header-includes:
  \usepackage{newunicodechar}
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{geometry}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \floatplacement{figure}{H}
  \usepackage{array}
  \usepackage{threeparttable}
  \usepackage{longtable}
  \usepackage{lscape}
  \usepackage{pdfpages}
output: 
  rticles::nejm_article:
    includes:
      in_header: ./preamble.tex
    latex_engine: lualatex
    keep_tex: true
    fig_caption: yes
    md_extensions: -autolink_bare_uris
geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"
bibliography: ./covid.bib
csl: ./plos.csl
always_allow_html: true
urlcolor: blue
linenumbers: false
linkcolor: cyan
---

@@

datadir <- "data/2021-03-16/"
load(file=paste0(datadir, "cc.all.RData"))
# with(cc.all[CASE==0 & SPECIMENDATE > as.Date("2021-02-28")], paste.rowpercent(table(dm.type, vax14.dose>0)))

@@

p.shieldvax <-
    ggplot(data=shield.bydate, aes(x=date, y=prop.vax)) +
    geom_line() +
    scale_y_continuous(breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1), limits=c(0, 1),
                       expand=c(0, 0)) + 
    scale_x_date(breaks = seq.Date(from = as.Date("2020-12-15"),
                                   to =  as.Date("2021-03-02"), by = "week"),
                 labels=gsub("^0", "", 
                             format.Date(seq.Date(from = as.Date("2020-12-15"),
                                                  to =  as.Date("2021-03-02"),
                                                  by = "week"),
                                         "%d %b")
                             ),
                 limits=c(as.Date("2020-12-15"), as.Date("2021-03-02"))) +
        labs(x=paste0("Presentation date: mid-point of 7-day window"),
         y="Proportion vaccinated among controls eligible for shielding")

keep.vars <- c("SPECIMENDATE", "CASE", "discharge15to56", "prob.hcai", "care.home", "hh.over18", "listedgr3", "shield.any", "shield.group", "dm.type3",
               listed.conditions, "shield.group", "vax14", "vax14.dose", "stratum")
cc.vax <- na.omit(cc.severe[SPECIMENDATE > as.Date("2020-12-01"), ..keep.vars])
cc.vax[, listedgr5 := listedgr3]
cc.vax[listedgr3=="Moderate risk condition" & dm.type != "Not diabetic",
       listedgr5 := dm.type3]
cc.vax[, listedgr5 := factor(as.character(listedgr5),
                     levels=c("No risk condition",
                              "Type 1 diabetes", "Type 2 diabetes",
                              "Moderate risk condition",
                              "Eligible for shielding"))]

@@

keep.hosp.vars <- c("SPECIMENDATE", "CASE", "discharge15to56", "care.home", "shieldelig.group", 
                    "vax14", "stratum")
cc.hosp.vax <- na.omit(cc.all[(casegroup=="A" | casegroup=="B") &
                              SPECIMENDATE > as.Date("2020-12-01"), ..keep.hosp.vars])
rm(cc.all)

@@

1836 words excluding Appendix

@@

# Abstract
- This study examines the efficacy of vaccination in reducing the risk of severe COVID-19 among those designated as clinically extremely vulnerable and thus eligible for shielding in Scotland.
- This analysis is based on data from 1 December 1st 2020 to 16 March 2021 and therefore covers the initial dose of vaccine with very few second doses in the shielded having been administered as yet  in this period. 
- Efficacy of a single vaccination dose in protecting against severe COVID-19 was found to be as high or higher in those eligible for shielding, considered as a whole,  as in those without risk conditions.   
- Examination of  vaccine efficacy among those eligible for shielding broken down by the reason for shielding showed that for  solid organ transplant recipients the number of cases was too small for the efficacy of vaccination to be estimated.  For all other groups there was a statistically significant reduction in hospitalisation for COVID-19 associated with  vaccination and the magnitude of this reduction was similar to that in those who were not eligible for shielding. 
- Further data in larger populations are needed to evaluate vaccine efficacy in solid organ transplant recipients.  Until such data are available it should not be assumed that vaccination is efficacious in this group.

# Introduction
In Scotland up to March 2021 COVID-19 vaccination has been with either the ChAdOx1 nCoV-19 (Astra-Zeneca) or BNT162b2 mRNA (Pfizer-BioNtech) vaccines.  A primary-care based observational study up to 15 Feb 2021 found maximal efficacy of 70% for ChAdOx1 nCoV-19 and 85% for BNT162b2 mRNA at 28 days post-vaccination [@vasileiou_effectiveness_2021].  However the efficacy of COVID-19 vaccination among those designated as clinically extremely vulnerable is still uncertain; an official [letter](https://www.gov.scot/publications/coronavirus-covid-19-shielding-letters/) sent on 23 March 2021 to those on this list stated that "we do not yet have evidence of exactly how effective [the vaccine] is for the people on the shielding list."   This is relevant to whether individuals in this category should be advised to remain in shielding as population-wide restrictions are lifted.  This study examines the efficacy of vaccination in reducing the risk of severe COVID-19 among those designated as clinically extremely vulnerable in Scotland.

# Methods 
This analysis is based on the REACT-SCOT matched case-control study, established at the beginning of the epidemic to explore risk factors for severe COVID-19 in the population of Scotland.  The design has been described in detail previously [@mckeigue_rapid_2020]. In brief, for every incident case of COVID-19 in the population ten controls matched to one-year age, sex and primary care practice were selected using the Community Health Index database.  COVID-19 cases are those with a positive nucleic acid test, or a hospital admission of death with COVID-19 codes. The dataset is refreshed regularly and is linked to the vaccination database and to the regularly updated dataset of all individuals deemed eligible for the shielding programme.  This analysis is based on all cases presenting between 1 December 1st 2020 to 16 March 2021.

We used this matched case-control design to take advantage of data linkages already established. 
As previously, to minimise ascertainment bias we focused the analysis on  severe COVID-19,  defined as entry to critical care within 28 days of presentation or  fatal outcome (any death within 28 days of a positive test or any death for which COVID-19 was coded as underlying cause).   As described previously, cases and controls were grouped into one of three categories: no risk condition, risk condition designated previously by Public Health Scotland as moderate risk, eligible for shielding.  For further analyses, the shielding category was subdivided as described previously into six categories: solid organ transplant, specific cancers, severe respiratory conditions, rare conditions, on immunosuppressants, additional conditions.

Vaccination status in cases was coded as having had at least one dose of any vaccine at least 14 days before presentation date. This assigns cases (and their matched controls) who were vaccinated less than 14 days prior to the case date to the unexposed category.  The numbers of cases and controls were too few too allow further subdivision of time since vaccination.

The effect of vaccination in each of the clinical vulnerability categories was estimated in a conditional logistic regression model fitted to the full dataset, specifying effects $\beta_{R2}, \dots, \beta_{RJ}$ for the log rate ratio associated with risk categories 2 to $J$ ($\beta_{R1} = 0$ for the reference category $J=1$), and "nested" effects $\beta_{V1}, \dots, \beta_{VJ}$ for the log rate ratio associated with vaccination in each of the $J$ risk ategories.  The  incidence density sampling, design controls not only for the matching factors of age, sex and general practice but also for calendar time which would otherwise confound the association between vaccination and disease. Because severe COVID-19 is strongly associated with care home residence and care home residents were a priority group for vaccination, the model includes the log rate ratio $\beta_H$ associated with care home residence. <!---For the $i$th individual in a matched set whose care home residence status is $H_i$, is in risk category $j$ and has vaccination status $V_i$, the linear predictor $\Theta_i$ is $\beta_H H_i + \beta_{Rj} + \beta_{Vj} V_{i}$.  The conditional likelihood for a matched set where individual $c$ is the case is $\exp{\left( \Theta_c \right) } / \sum_i{\exp{\left( \Theta_i \right) }}$. ---> The efficacy of vaccination in the $j$th risk category is 1 minus the rate ratio $\exp{\left(\beta_{Vj} \right)}$. More explanation of the estimation of vaccine  efficacy in a matched case-control design are given in the Statistical Note in the Appendix.

@@

# Results
Figure \ref{fig:plotvax} shows the  proportion vaccinated at least 14 days before among those eligible for shielding in the control group, by presentation date.  This proportion in  controls  will be a close estimate of the  overall  vaccination rates in the total shielding population. From this we can infer that by 9 February 2021 about half the shielded population were vaccinated, rising to over 85% by early March.

```{r plotvax, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.width=4, fig.asp=0.8, fig.cap="\\label{fig:plotvax}Proportion vaccinated at least 14 days before among those eligible for shielding in the control group, by presentation date"}

@@

Supplementary Table \ref{tab:freqsshieldgroups} shows  the vaccination status of  severe cases and  their controls  by risk stratum.  Supplementary Table \ref{tab:freqshosp} shows the same breakdown for hospitalised cases and controls.   These tables show that the numbers of cases among solid organ transplant recipients, are too small for rate ratios to be estimated reliably.

The rate ratios for severe COVID-19 associated with vaccination were `r format.estci(table.riskgroups$rateratio[4])` in those with no risk condition, `r format.estci(table.riskgroups$rateratio[5])` in those with moderate risk conditions and `r format.estci(table.riskgroups$rateratio[6])` in those eligible for shielding.  With adjustment for having been in hospital for more than 7 days before presentation and for any hospital discharge from 15 to 56 days before presentation, the corresponding rate ratios were `r format.estci(table.adjhosp$rateratio[6])`, `r format.estci(table.adjhosp$rateratio[7])` and `r format.estci(table.adjhosp$rateratio[8])`.  
Table \ref{tab:shieldgroups} gives a breakdown by shielding group.  Point estimates for the rate ratio associated with vaccination were lower in all risk groups than in those without risk conditions,  with the exception of solid organ transplant recipients,  those  with rare diseases and those on immunosuppressants in whom the numbers of cases and were small (Supplementary Table \ref{tab:freqsshieldgroups}) and confidence intervals were wide.   Among those in the "moderate risk conditions" category who had diabetes, the rate ratios for severe COVID-19 associated with vaccination were `r format.estci(table.dmriskgroups$rateratio[2])` and `r format.estci(table.dmriskgroups$rateratio[3])` in those with Type 1 and Type 2 diabetes respectively.

@@

Table \ref{tab:shieldelig} shows the analysis extended to include all hospitalized cases, for which the numbers of cases are larger.   This shows significant efficacy in all groups other than solid organ transplant recipients but even with the broader case definition the numbers of cases and controls were too small  to evaluate efficacy. (Supplementary Table \ref{tab:freqshosp}).

@@

A study in a larger population is needed to estimate efficacy in solid organ transplant recipients. Such an analysis would lay the basis for advising people in this group about their ongoing risk level, and for assessing the need for development of alternatives to vaccines such as synthetic antibodies that are now being evaluated in clinical trials [@eli_lilly_and_company_phase_2021]

The Vaccine Green Book recommends that the second dose of all  COVID-19 vaccines should be routinely scheduled between 4 and 12 weeks after the first dose [@public_health_england_covid-19_2021].  Although earlier rather than later second dosing within that interval might more quickly reduce the residual risk among those in shielding,   this is not necessarily the optimal strategy as the JCVI have noted that "With most vaccines an extended interval between the prime and booster doses leads to a better immune response to the booster dose." [@joint_committee_on_vaccination_and_immunisation_optimising_2021]

## Competing interest
The authors declare no competing interests

\beginsupplement
\onehalfspacing
\newpage

# Appendix

```{r freqsshieldgroups, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(freqs.hosp, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="freqshosp",
             row.names=TRUE,
  		 col.names=c("Unvaccinated", "Vaccinated", "Unvaccinated", "Vaccinated"),
             align=c("l", "r", "r", "r", "r"),
             caption="Numbers of controls and cases of hospitalized or fatal COVID-19 by risk group and vaccination status") %>%
    add_header_above(c(" "=1, "Controls"=2, "Cases"=2)) %>% 
    add_footnote(label=c("Presentation dates from 1 December 2020 to 16 March 2021",
                         "Vaccine status coded as 1 if at least one dose at least 14 days before, 0 otherwise"), 
                 notation="none")  %>%
    column_spec(2, width="2cm") %>%
    column_spec(3, width="2cm") %>%
    column_spec(4, width="2cm") %>%
    column_spec(5, width="2cm") %>%
    column_spec(6, width="1cm") %>%
    kable_styling(latex_options=c("hold_position"))

```{r mh.dates, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(mh.dates[1:17, ], 
             escape=FALSE, 
             booktabs=TRUE,
			 label="mhdates",
             row.names=FALSE,
             align=c("l", "r", "r", "r", "r"),
             digits=1,
             caption="Vaccinated and unvaccinated severe cases among those eligible for shielding and not resident in care homes: observed and expected numbers by week") %>%
    add_header_above(c(" "=1, "Vaccinated cases"=2, "Unvaccinated cases"=2)) %>% 
    add_footnote(label=c("Expected numbers (under the null) are calculated in each stratum",
                         "Vaccine status coded as 1 if at least one dose at least 14 days before, 0 otherwise"), 
                 notation="none")  %>%
    kable_styling(latex_options=c("hold_position"))

## Statistical note
Tables \ref{tab:freqsshieldgroups} and  \ref{tab:freqshosp}  show the the vaccination status of  distribution of cases and controls by shielding stratum  to illustrate that for solid organ transplant recipients the numbers of cases are too few to estimate vaccine efficacy reliably.   
We have used a matched case-control design, in which each incident case was matched to up to 10 controls of the same age (one-year band), sex and GP practice who were alive on the day of presentation of the case, forming a matched set.  Analyses of these matched sets are thus  inherently stratified by day.  Most matched sets have 1 case and up to 10 controls, but where two cases with the same age, sex and GP practice present on the same day the matched set would have 2 cases and up to 20 controls.

<!---
Each matched set can be represented as a 2 \ensuremath{\times} 2 table, in which the numbers in each cell are traditionally labelled $a, b, c, d$:

 | | Case | Control | All |
 | --- | --- | --- | --- |
 Vaccinated | $a$ | $b$ | $V_1$ |
 Unvaccinated | $c$ | $d$ | $V_0$ |
 All | $C_1$ | $C_0$ | $N$ |
--->
In brief vaccine efficacy is estimated as the most likely value of the odds ratio associated with vaccination given the observed values in each matched set, conditional on the marginal totals (total cases, total controls, total vaccinated, total unvaccinated).  The log-likelihood of the odds ratio given the data is summed across all sets: matched sets in which all are vaccinated or none are vaccinated do not contribute information.  When estimating the efficacy within a risk group such as a shielding category, only cases and controls in that risk group are retained in the matched sets

It follows from the above that crude tables of the vaccination status of cases and controls in each risk group cannot be used to estimate the efficacy.  This is especially true given that vaccination rates were rising sharply over the study period and rates of severe disease were falling.  Table \ref{tab:mhdates} may give the reader a more intuitive sense of how the information for computing vaccine efficacy in the shielded accumulated over the study period.  It shows the weekly observed split of vaccinated versus unvaccinated cases among those in shielding versus the expected split conditional upon the marginal totals.  To keep the illustration simple, this table is restricted to the 778 cases eligible for shielding who  were not resident in care homes so there is no covariate adjustment.  The variance of the expected can be interpreted as the information about vaccine efficacy. As shown fewer of the cases were vaccinated and more were unvaccinated than expected if vaccination had no impact on disease rates.

>>>>

---
title: "Brief Report: efficacy of COVID-19 vaccination in individuals designated as clinically extremely vulnerable in Scotland"
author:
  - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
    affiliation: HPS
  - name: David A McAllister\textsuperscript{\getAff{Glasgow}}  
    affiliation: HPS 
  - name: Jen Bishop
    affiliation: HPS
  - name: Sharon Hutchinson\textsuperscript{\getAff{GCU}} 
    affiliation: HPS
  - name: Chris Robertson\textsuperscript{\getAff{Strathclyde}}   
    affiliation: HPS
  - name: Nazir Lone
    affiliation: Usher
  - name: Jim McMenamin 
    affiliation: HPS
  - name: David Goldberg
    affiliation: HPS
  - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
    corresponding: yes
    email: helen.colhoun@igmm.ed.ac.uk
    affiliation: HPS
address: 
- code: Usher
  address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care  
- code: HPS
  address: Public Health Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
- code: Glasgow
  address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
- code: GCU
  address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
- code: Strathclyde
  address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
- code: IGMM
  address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - AXA Chair in Medical Informatics and Epidemiology

header-includes:
  \usepackage{newunicodechar}
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{geometry}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \floatplacement{figure}{H}
  \usepackage{array}
  \usepackage{threeparttable}
  \usepackage{longtable}
output: 
  rticles::nejm_article:
    includes:
      in_header: ./preamble.tex
    latex_engine: lualatex
    keep_tex: true
    fig_caption: yes
    md_extensions: -autolink_bare_uris
geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"
bibliography: ./covid.bib
csl: ./plos.csl
always_allow_html: true
urlcolor: blue
linenumbers: false
linkcolor: cyan
---

@@

datadir <- "data/2021-03-16/"
load(file=paste0(datadir, "cc.all.RData"))

@@

p.shieldvax <-
    ggplot(data=shield.bydate, aes(x=date, y=prop.vax)) +
    geom_line() +
    scale_y_continuous(breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1), limits=c(0, 1),
                       expand=c(0, 0)) + 
    scale_x_date(breaks = seq.Date(from = as.Date("2020-12-15"),
                                   to = as.Date("2021-03-02"), by = "week"),
                 labels=gsub("^0", "", 
                             format.Date(seq.Date(from = as.Date("2020-12-15"),
                                                  to =  as.Date("2021-03-02"),
                                                  by = "week"),
                                         "%d %b")
                             ),
                 limits=c(as.Date("2020-12-15"), as.Date("2021-03-02"))) +
        labs(x=paste0("Sampling date: mid-point of 7-day window"),
         y="Proportion vaccinated among controls eligible for shielding")

keep.vars <- c("SPECIMENDATE", "CASE", "discharge15to56", "prob.hcai", "care.home", "hh.over18", "listedgr3", "shield.any", "shield.group", "dm.type3", "fatalcase",
               listed.conditions, "shield.group", "vax14", "vax14.dose", "stratum")
cc.vax <- na.omit(cc.severe[SPECIMENDATE > as.Date("2020-12-01"), ..keep.vars])
cc.vax[, listedgr5 := listedgr3]
cc.vax[listedgr3=="Moderate risk condition" & dm.type != "Not diabetic",
       listedgr5 := dm.type3]
cc.vax[, listedgr5 := factor(as.character(listedgr5),
                     levels=c("No risk condition",
                              "Type 1 diabetes", "Type 2 diabetes",
                              "Moderate risk condition",
                              "Eligible for shielding"))]

@@

keep.hosp.vars <- c("SPECIMENDATE", "CASE", "discharge15to56", "care.home", "shieldelig.group", 
                    "vax14", "stratum")
cc.hosp.vax <- na.omit(cc.all[(casegroup=="A" | casegroup=="B") &
                              SPECIMENDATE > as.Date("2020-12-01"), ..keep.hosp.vars])
#rm(cc.all)

@@

2201 words, abstract 270 words

@@

# Abstract
* _Objectives_ - The objective of this study was to estimate the efficacy of vaccination in reducing the risk of severe COVID-19 among those designated as clinically extremely vulnerable in Scotland.    
* _Methods_ -  In a matched case-control design (REACT-SCOT), all `r nrow(cc.all[CASE==1 & SPECIMENDATE >= as.Date("2020-12-01")])` cases of COVID-19 in Scotland diagnosed from 1 December 2020 to 16 March 2021 were matched for age, sex and primary care practice to `r nrow(cc.all[CASE==0 & SPECIMENDATE >= as.Date("2020-12-01")])` controls from the general population.  This dataset was linked to vaccination records and to the list of those identified as eligible for shielding.  Severe COVID-19 was  defined by entry to critical care or fatal outcome. Rate ratios were estimated by conditional logistic regression.  
* _Results_ -  The rate ratio for severe COVID-19 associated with vaccination at least 14 days before was `r format.estci(table.riskgroups$rateratio[6])` in those eligible for shielding, compared with `r format.estci(table.shieldgroups[9, 2])` in those without risk conditions.  In a breakdown by shielding condition, the rate ratio for the broader category of hospitalization or fatal outcome ranged from `r format.estci(table.hosp[10, 2])` in those with specific cancers to `r format.estci(table.hosp[9, 2])` in solid organ transplants recipients, and  `r format.estci(table.hosp[13, 2])` in others on immunosuppressants (excluding solid organ transplant recipients)  compared with `r format.estci(table.hosp[8, 2])` in those ineligible for shielding.  
* _Conclusions_ -  These results, obtained when very few second doses of vaccine had been administered, are reassuring with respect to efficacy in clinically vulnerable individuals including immunocompromised individuals, but studies in larger populations are needed to estimate vaccine efficacy in solid organ transplant recipients.

# Introduction
In Scotland up to March 2021 only the ChAdOx1 nCoV-19 (Astra-Zeneca) and BNT162b2 mRNA (Pfizer-BioNtech) vaccines were used.  A primary-care based observational study up to 15 Feb 2021 found maximal efficacy of 70% for ChAdOx1 nCoV-19 and 85% for BNT162b2 mRNA at 28 days post-vaccination [@vasileiou_effectiveness_2021].  However the efficacy of COVID-19 vaccination among those designated as clinically extremely vulnerable is still uncertain; an official [letter](https://www.gov.scot/publications/coronavirus-covid-19-shielding-letters/) sent on 23 March 2021 to those on this list stated that "we do not yet have evidence of exactly how effective [the vaccine] is for the people on the shielding list."   This is relevant to whether individuals in this category should be advised to remain in shielding as population-wide restrictions are lifted.  This study examines the efficacy of vaccination in reducing the risk of severe COVID-19 among those designated as clinically extremely vulnerable in Scotland.

# Methods 
This analysis is based on the REACT-SCOT matched case-control study, established at the beginning of the epidemic to investigate risk factors for severe COVID-19 in the population of Scotland.  We used this study to take advantage of data linkages already established. 
The design has been described in detail previously [@mckeigue_rapid_2020]. In brief, for every incident case of COVID-19 in the population ten controls matched to one-year age, sex and primary care practice and alive on the day of presentation of the case were selected using the Community Health Index database.  COVID-19 cases are those with a positive nucleic acid test, or a hospital admission of death with COVID-19 codes. The dataset is refreshed regularly and is linked to the vaccination database and to the regularly updated dataset of all individuals deemed eligible for the shielding programme.  This analysis is based on all cases presenting from 1 December 2020 to 16 March 2021.

As previously, to minimise ascertainment bias we focused the analysis on  severe COVID-19,  defined as entry to critical care within 28 days of presentation or  fatal outcome (any death within 28 days of a positive test or any death for which COVID-19 was coded as underlying cause).   As previously, cases and controls were grouped into one of three categories: no risk condition, risk condition designated previously by Public Health Scotland as moderate risk, eligible for shielding.  For further analyses, the shielding category was subdivided as described previously into six categories: solid organ transplant, specific cancers, severe respiratory conditions, rare conditions, on immunosuppressants, additional conditions.

Vaccination status in cases was coded as having had at least one dose of any vaccine at least 14 days before presentation date. This assigns cases (and their matched controls) who were vaccinated less than 14 days prior to the case date to the unexposed category.  The numbers of cases and controls were too few to allow further subdivision of time since vaccination.

The effect of vaccination in each of the clinical vulnerability categories was estimated in a conditional logistic regression model fitted to the full dataset, specifying effects $\beta_{R2}, \dots, \beta_{RJ}$ for the log rate ratio associated with risk categories 2 to $J$ ($\beta_{R1} = 0$ for the reference category $J=1$), and "nested" effects $\beta_{V1}, \dots, \beta_{VJ}$ for the log rate ratio associated with vaccination in each of the $J$ risk categories. The strata in this conditional logistic regression model are matched sets: typically one case and up to ten controls, but where two cases with the same age, sex and GP practice present on the same day the matched set comprises two cases and up to 20 controls.  When estimating the efficacy within a risk group such as a shielding category, only cases and controls in that risk group are retained in the matched sets.   Matched sets in which all are vaccinated or none are vaccinated do not contribute to the conditional odds ratio.

With this incidence density sampling design, the conditional odds ratio is the rate ratio and the efficacy of vaccination is 1 minus the rate ratio.  The unconditional odds ratio calculated from summary tables of the vaccination status of cases and controls in each risk group cannot be used to estimate the rate ratio.

The design controls not only for the matching factors of age, sex and general practice but also for calendar time which would otherwise strongly confound the association between vaccination and disease. Because severe COVID-19 is strongly associated with care home residence and care home residents were a priority group for vaccination, the model includes the log rate ratio $\beta_H$ associated with care home residence.

@@

# Results
Figure \ref{fig:plotvax} shows the  proportion vaccinated at least 14 days before among those eligible for shielding in the control group, by seven-day sliding window of the date on which the controls were sampled. From this we can estimate that by 9 February 2021 about half the shielded population were vaccinated, rising to over 85% by early March.

Table \ref{tab:freqsshieldgroups} shows  the vaccination status of  severe cases and  their controls  by risk stratum.   `r cc.vax[CASE==1 & fatalcase, .N, by=listedgr3][2, 2]` of the `r cc.vax[CASE==1, .N, by=listedgr3][2, 2]` severe cases among those eligible for shielding were fatal.  For comparison, `r cc.vax[CASE==1 & fatalcase, .N, by=listedgr3][3, 2]` of the `r cc.vax[CASE==1, .N, by=listedgr3][3, 2]` severe cases among with no risk condition were fatal.

Table \ref{tab:freqshosp} shows the same breakdown for hospitalised cases and controls. 
The rate ratios for severe COVID-19 associated with vaccination were `r format.estci(table.shieldgroups[9, 2])` in those with no risk condition, `r format.estci(table.shieldgroups[10, 2])` in those with moderate risk conditions and `r format.estci(table.riskgroups$rateratio[6])` in those eligible for shielding.  With adjustment for having been in hospital for more than 7 days before presentation and for any hospital discharge from 15 to 56 days before presentation, the corresponding rate ratios were `r format.estci(table.adjhosp$rateratio[6])`, `r format.estci(table.adjhosp$rateratio[7])` and `r format.estci(table.adjhosp$rateratio[8])`.

Table \ref{tab:shieldgroups} gives a breakdown by shielding group.  Point estimates for the rate ratio associated with vaccination were lower in all risk groups than in those without risk conditions,  with the exception of solid organ transplant recipients,  those  with rare diseases and those on immunosuppressants in whom the numbers of cases were small (Table \ref{tab:freqsshieldgroups}) and confidence intervals were wide.   Among those in the "moderate risk conditions" category who had diabetes, the rate ratios for severe COVID-19 associated with vaccination were `r format.estci(table.dmriskgroups$rateratio[2])` and `r format.estci(table.dmriskgroups$rateratio[3])` in those with Type 1 and Type 2 diabetes respectively.

Table \ref{tab:shieldelig} shows the analysis extended to include all hospitalized cases, for which the numbers of cases are larger.   This shows evidence for efficacy in all groups except solid organ transplant recipients in whom even with this broader case definition the numbers of cases and controls were too small for efficacy to be estimated reliably.

As rates of severe disease in Scotland have now fallen to very low levels, the efficacy of two doses compared with one cannot be estimated accurately. The Vaccine Green Book recommends that the second dose of all  COVID-19 vaccines should be routinely scheduled between 4 and 12 weeks after the first dose [@public_health_england_covid-19_2021].  Although earlier rather than later second dosing within that interval might more quickly reduce the residual risk among those in shielding,   this is not necessarily the optimal strategy as the JCVI have noted that "With most vaccines an extended interval between the prime and booster doses leads to a better immune response to the booster dose." [@joint_committee_on_vaccination_and_immunisation_optimising_2021].

A study in a larger population is needed to estimate efficacy in solid organ transplant recipients. Such an analysis would lay the basis for advising people in this group about their ongoing risk level, and for assessing the need for development of alternatives to vaccines such as synthetic antibodies that are now being evaluated in clinical trials  [@eli_lilly_and_company_phase_2021].

## Competing interests
The authors declare no competing interests

```{r plotvax, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.width=4, fig.asp=0.8, fig.cap="\\label{fig:plotvax}Proportion vaccinated at least 14 days before among those eligible for shielding in the control group, by 7-day window of date on which they were sampled (presentation date of the cases they were matched to)"}

@@

```{r freqsshieldgroups, echo=FALSE, message=FALSE}

knitr::kable(freqs.hosp, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="freqshosp",
             row.names=TRUE,
  		 col.names=c("Unvaccinated", "Vaccinated", "Unvaccinated", "Vaccinated"),
             align=c("l", "r", "r", "r", "r"),
             caption="Numbers of controls and cases of hospitalized or fatal COVID-19 by risk group and vaccination status") %>%
    add_header_above(c(" "=1, "Controls"=2, "Cases"=2)) %>% 
    add_footnote(label=c("Presentation dates from 1 December 2020 to 16 March 2021",
                         "Vaccine status coded as 1 if at least one dose at least 14 days before, 0 otherwise"), 
                 notation="none")  %>%
    column_spec(2, width="2cm") %>%
    column_spec(3, width="2cm") %>%
    column_spec(4, width="2cm") %>%
    column_spec(5, width="2cm") %>%
    column_spec(6, width="1cm") %>%
    kable_styling(latex_options=c("HOLD_position"))

@@

@@
